---
title: "NFL Data Bowl"
author: "Konner Macias"
date: "December 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read the data
```{r}
library(tidyverse)
file.tracking <- "https://raw.githubusercontent.com/nfl-football-ops/Big-Data-Bowl/master/Data/tracking_gameId_2017090700.csv"
tracking.example <- read_csv(file.tracking)

file.game <- "https://raw.githubusercontent.com/nfl-football-ops/Big-Data-Bowl/master/Data/games.csv"
games.sum <- read_csv(file.game) 

file.plays <- "https://raw.githubusercontent.com/nfl-football-ops/Big-Data-Bowl/master/Data/plays.csv"
plays.sum <- read_csv(file.plays) 

tracking.example.merged <- tracking.example %>% inner_join(games.sum) %>% inner_join(plays.sum) 

example.play <- tracking.example.merged %>% filter(playId == 2756)

example.play %>% select(playDescription) %>% slice(1)
```

# Animate data
```{r}
library(gganimate)
library(cowplot)

## General field boundaries
xmin <- 0
xmax <- 160/3
hash.right <- 38.35
hash.left <- 12
hash.width <- 3.3


## Specific boundaries for a given play
ymin <- max(round(min(example.play$x, na.rm = TRUE) - 10, -1), 0)
ymax <- min(round(max(example.play$x, na.rm = TRUE) + 10, -1), 120)
df.hash <- expand.grid(x = c(0, 23.36667, 29.96667, xmax), y = (10:110))
df.hash <- df.hash %>% filter(!(floor(y %% 5) == 0))
df.hash <- df.hash %>% filter(y < ymax, y > ymin)
```

Animation Function
```{r}
animate.play <- function(play) {
  ggplot() +
  geom_point(data = play, aes(x = (xmax-y), y = x, 
                                 colour = team, group = nflId, pch = team, size = team)) + 
  geom_text(data = play, aes(x = (xmax-y), y = x, label = jerseyNumber), colour = "white", 
            vjust = 0.36, size = 3.5) + 
  scale_size_manual(values = c(6, 4, 6), guide = FALSE) + 
  scale_shape_manual(values = c(19, 16, 19), guide = FALSE) +
  scale_colour_manual(values = c("#e31837", "#654321", "#002244"), guide = FALSE) + 
  annotate("text", x = df.hash$x[df.hash$x < 55/2], 
           y = df.hash$y[df.hash$x < 55/2], label = "_", hjust = 0, vjust = -0.2) + 
  annotate("text", x = df.hash$x[df.hash$x > 55/2], 
           y = df.hash$y[df.hash$x > 55/2], label = "_", hjust = 1, vjust = -0.2) + 
  annotate("segment", x = xmin, 
           y = seq(max(10, ymin), min(ymax, 110), by = 5), 
           xend =  xmax, 
           yend = seq(max(10, ymin), min(ymax, 110), by = 5)) + 
  annotate("text", x = rep(hash.left, 11), y = seq(10, 110, by = 10), 
                    label = c("G   ", seq(10, 50, by = 10), rev(seq(10, 40, by = 10)), "   G"), 
                    angle = 270, size = 4) + 
  annotate("text", x = rep((xmax - hash.left), 11), y = seq(10, 110, by = 10), 
           label = c("   G", seq(10, 50, by = 10), rev(seq(10, 40, by = 10)), "G   "), 
           angle = 90, size = 4) + 
  annotate("segment", x = c(xmin, xmin, xmax, xmax), 
           y = c(ymin, ymax, ymax, ymin), 
           xend = c(xmin, xmax, xmax, xmin), 
           yend = c(ymax, ymax, ymin, ymin), colour = "black") + 
  ylim(ymin, ymax) + 
  coord_fixed() +  
  theme_nothing() + 
  transition_time(frame.id)  +
  ease_aes('linear') + 
  NULL
}
```

Now animate
```{r}
simPlay <- function(df, play.id) {
  play <- df %>% filter(playId == play.id)
  play.length <- length(unique(play$frame.id))
  animate(animate.play(play), fps = 10, nframe = play.length)
}
```


# Real work
Next Goal, understand play routes

Data Frame Goal:
- Contains all successful throwing plays with Gronk present on field
```{r}
# Gather game IDs of all Patriots games
pat.game.ids <- games.sum[games.sum$homeTeamAbbr == "NE" | games.sum$visitorTeamAbbr == "NE",]$gameId

# create patriot game gathering fn
getPatGame <- function(game.id) {
  url <- "https://raw.githubusercontent.com/nfl-football-ops/Big-Data-Bowl/master/Data/tracking_gameId_"
  tail <- ".csv"
  file.tracking <- paste0(url, game.id, tail)
  game.inst <- read_csv(file.tracking)
  game.inst.merged <- game.inst %>% inner_join(games.sum) %>% inner_join(plays.sum) 
  return(game.inst.merged)
}
```
Practice on first instance
```{r}
pg1 <- getPatGame(pat.game.ids[1])
head(pg1)
```

Now breakup dataframe to include only plays with gronk on field
```{r}
gr.play.ids <- pg1[pg1$displayName == "Rob Gronkowski",]$playId
gr.plays1 <- pg1[pg1$playId %in% unique(gr.play.ids),]
head(gr.plays1)
```

```{r}
#simPlay(gr.plays1, 629)
```



```{r}
# create subframe of just passes
pplay.ids <- gr.plays1[grepl("pass",gr.plays1$playDescription),]$playId
suc.pplay.ids <- gr.plays1[grepl("pass",gr.plays1$playDescription) & 
                             !grepl("incomplete",gr.plays1$playDescription),]$playId
gr.pplays1 <- gr.plays1[gr.plays1$playId %in% unique(pplay.ids),]
gr.spplays1 <- gr.plays1[gr.plays1$playId %in% unique(suc.pplay.ids),]
```

Look at successful passes
```{r}
#unique(gr.spplays1$playId)
```

```{r}
#simPlay(gr.spplays1,3846)
```

Notes:
playId == 449 : Gronk brought help (56,38) away from intended receiver
playId == 629 : Gronk brought main defender (49) and help (56,29) away from intended receiver
playId == 1049 : Gronk included in sweep to move group of defenders off
playId == 3528 : Gronk helps by pulling defenders off for targeted receiver to score
playId == 3776 : Gronk moves back behind line of scrimmage to lead defenders in to distract
- We need to tag these plays as successful receiver route!
HOW


```{r}
file.players <- "https://raw.githubusercontent.com/nfl-football-ops/Big-Data-Bowl/master/Data/players.csv"
players.sum <- read_csv(file.players) 
players.sum <- players.sum[,c(1,4)]
```
Now add position
```{r}
# hard code
gr.spplays1$Position <- NA
for (nfl.id in gr.spplays1$nflId) {
  gr.spplays1[!is.na(gr.spplays1$nflId) & gr.spplays1$nflId == nfl.id,]$Position <- 
    players.sum[players.sum$nflId == nfl.id,]$PositionAbbr
}
```

## Analyzing movement

Use direction. Use t - (t-1) for x and y to determine for/back/in/out. 

 x120
y53.3  y0
______
|    |
|    |
|    |
|    |
______
  x0

Team Direction: DOWN|UP (My perspective)
xt - xt_1 > 0 : Forward|Backward
yt - yt_1 > 0 : Right/Left | Left/Right
Dir:
Split into 8 segments:
330-30: W, 30-60: NW, 60-120: N,
120-150: NE, 150-210: E, 210-240: SE,
240-300: S, 300-330: SW

```{r}
giveDir <- function(dir) {
  ifelse (dir > 330 | dir < 30, "W",
          ifelse(dir >= 30 & dir < 60, "NW",
          ifelse(dir >= 60 & dir < 120, "N",
          ifelse(dir >= 120 & dir < 150, "NE",
          ifelse(dir >= 150 & dir < 210, "E",
          ifelse(dir >= 210 & dir < 240, "SE",
          ifelse(dir >= 240 & dir < 300, "S", "SW")))))))
}
```
Create new column
```{r}
gr.spplays1$dirFac <- lapply(gr.spplays1$dir, giveDir)
gr.spplays1$dirFac <- as.factor(unlist(gr.spplays1$dirFac))
```

Next, define team direction.
If home team has higher x than away, then they are heading south.
Else north. Since this data frame is only NE on offense.

```{r}
# set offensive direction
# first set all to NA
gr.spplays1$OffDir <- NA
for (play.id in unique(gr.spplays1$playId)) {
  df <- gr.spplays1[gr.spplays1$playId==play.id,]
  away.ind <- min(which(df$team == "away"))
  home.ind <- min(which(df$team == "home"))
  # compare the x of both
  gr.spplays1[gr.spplays1$playId==play.id,]$OffDir <- 
    ifelse(df$x[away.ind] > df$x[home.ind], "North", "South")
}
```



# Defining Routes
Different receiver routes: Flat, Slant, Comeback, Curl, Out, In/Dig, Corner, Post, Fly

Target the receivers and tightends.

Only check for after the snap, and before the catch.
If OffDir == South -> check for South. If y > 26.65 check East. If y < 26.65 check west.
else -> Check for North if y > 26.65 check east. If y < 26.65 check west.


```{r}
InDigRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left
    ifelse("S" %in% df$dirFac & "E"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="E")), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "W"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="W")), T, F),
        F)
    
    # north left
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "E"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="E")),T,F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "W"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="W")),T,F),
        F)
  }
}
```

```{r}
OutRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left
    ifelse("S" %in% df$dirFac & "W"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="W")), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "E"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="E")), T, F),
        F)
    
    # north left
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "W"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="W")),T,F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "E"  %in% df$dirFac,
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="E")),T,F),
        F)
  }
}
```

Slant Route
- must contain only N and NE, no E and so on...
```{r}
SlantRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left
    ifelse("S" %in% df$dirFac & "SE" %in% df$dirFac & !("E" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="SE")), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "SW" %in% df$dirFac & !("W" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="SW")), T, F),
        F)
    
    # north left
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "NE" %in% df$dirFac & !("E" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="NE")),T,F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "NW" %in% df$dirFac & !("W" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="NW")),T,F),
        F)
  }
}
```

Curl Route: Go N -> NE -> E -> SE? -> S?

NOT WORKING 100% - too strict
ex: play 94, player 14 ran a curl route
```{r}
CurlRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left - turning counterclockwise
    ifelse("S" %in% df$dirFac & "SE" %in% df$dirFac & "E" %in% df$dirFac & "NE" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="S")) < min(which(df$dirFac=="SE")) &
                  (min(which(df$dirFac=="SE")) < min(which(df$dirFac=="E"))) &
                  (min(which(df$dirFac=="E")) < min(which(df$dirFac=="NE")))), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "SW" %in% df$dirFac & "W" %in% df$dirFac & "NW" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="S")) < min(which(df$dirFac=="SW")) &
                  (min(which(df$dirFac=="SW")) < min(which(df$dirFac=="W"))) &
                  (min(which(df$dirFac=="W")) < min(which(df$dirFac=="NW")))), T, F),
        F)
    
    # north left - turn clock wise
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "NE" %in% df$dirFac & "E" %in% df$dirFac & "SE" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="N")) < min(which(df$dirFac=="NE")) &
                  (min(which(df$dirFac=="NE")) < min(which(df$dirFac=="E"))) &
                  (min(which(df$dirFac=="E")) < min(which(df$dirFac=="SE")))), T, F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "NW" %in% df$dirFac & "W" %in% df$dirFac & "SW" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="N")) < min(which(df$dirFac=="NW")) &
                  (min(which(df$dirFac=="NW")) < min(which(df$dirFac=="W"))) &
                  (min(which(df$dirFac=="W")) < min(which(df$dirFac=="SW")))), T, F),
        F)
  }
}
```

Comeback route is curl but opposite way
```{r}
ComebackRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left - turning clockwise
    ifelse("S" %in% df$dirFac & "SW" %in% df$dirFac & "W" %in% df$dirFac & "NW" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="S")) < min(which(df$dirFac=="SW")) &
                  (min(which(df$dirFac=="SW")) < min(which(df$dirFac=="W"))) &
                  (min(which(df$dirFac=="W")) < min(which(df$dirFac=="NW")))), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "SE" %in% df$dirFac & "E" %in% df$dirFac & "NE" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="S")) < min(which(df$dirFac=="SE")) &
                  (min(which(df$dirFac=="SE")) < min(which(df$dirFac=="E"))) &
                  (min(which(df$dirFac=="E")) < min(which(df$dirFac=="NE")))), T, F),
        F)
    
    # north left - turn counter clockwise
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "NW" %in% df$dirFac & "W" %in% df$dirFac & "SW" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="N")) < min(which(df$dirFac=="NW")) &
                  (min(which(df$dirFac=="NW")) < min(which(df$dirFac=="W"))) &
                  (min(which(df$dirFac=="W")) < min(which(df$dirFac=="SW")))), T, F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "NE" %in% df$dirFac & "E" %in% df$dirFac & "SE" %in% df$dirFac,
        ifelse((min(which(df$dirFac=="N")) < min(which(df$dirFac=="NE")) &
                  (min(which(df$dirFac=="NE")) < min(which(df$dirFac=="E"))) &
                  (min(which(df$dirFac=="E")) < min(which(df$dirFac=="SE")))), T, F),
        F)
  }
}
```

```{r}
CornerRoute <- function(df, locType) {
  if (locType == "SL") {
    # south left
    ifelse("S" %in% df$dirFac & "SW" %in% df$dirFac & !("W" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="SW")), T, F),
        F)
    
    # south right
  } else if (locType == "SR") {
    ifelse("S" %in% df$dirFac & "SE" %in% df$dirFac & !("E" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="S")) < min(which(df$dirFac=="SE")), T, F),
        F)
    
    # north left
  } else if (locType == "NL") {
    ifelse("N" %in% df$dirFac & "NW" %in% df$dirFac & !("W" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="NW")),T,F),
        F)
    
    # north right
  } else if (locType == "NR") {
    ifelse("N" %in% df$dirFac & "NE" %in% df$dirFac & !("E" %in% df$dirFac),
        ifelse(min(which(df$dirFac=="N")) < min(which(df$dirFac=="NE")),T,F),
        F)
  }
}
```

```{r}
FlyRoute <- function(df, locType) {

  if (df$PassLength[1] < 15) return (FALSE)
  
  if (locType == "SL" | locType == "SR") {
    if ("S" %in% df$dirFac){
      fr.df <- as.data.frame(prop.table(table(df$dirFac))['S'])
      colnames(fr.df) <- "freq"
      return (ifelse (fr.df$freq[1] > 0.92, T, F)) ##go bruins
      } else {return (FALSE)}
  }

  if (locType == "NL" | locType == "NR") {
    if ("N" %in% df$dirFac){
      fr.df <- as.data.frame(prop.table(table(df$dirFac))['N'])
      colnames(fr.df) <- "freq"
      return (ifelse (fr.df$freq[1] > 0.92, T, F)) ##go bruins
      } else {return (FALSE)}
  }
  
}
```



```{r}
routeHandler <- function(df) {
  # make sure position is receiver
  if (df$Position[1] %in% c("WR","TE")) {
    # check which direction offense is heading
    if (df$OffDir[1] == "South") {
      # check left side of field
      if (df$y[1] > 26.65) {
        # determine route type
        if (InDigRoute(df,"SL")) return("In/Dig")
        if (OutRoute(df,"SL")) {
          if (df$PassLength[1] > 8) {return("Out")} else {return ("Flat")} ##8 clap go bruins
          }
        if (SlantRoute(df,"SL")) {
          if (df$PassLength[1] > 15) {return("Post")} else {return ("Slant")}
        }
        if (df$PassLength[1] > 15 & CornerRoute(df, "SL")) return ("Corner")
        if (SlantRoute(df,"SL")) return("Slant")
        if (CurlRoute(df,"SL")) return("Curl")
        if (ComebackRoute(df,"SL")) return("Comeback")
        if (FlyRoute(df, "SL")) return("Fly")
        
        
      } else {
        # player is on right side of field
        if (InDigRoute(df,"SR")) return("In/Dig")
        if (OutRoute(df,"SR")) {
          if (df$PassLength[1] > 8) {return("Out")} else {return ("Flat")} ##8 clap go bruins
          }
        if (SlantRoute(df,"SR")) {
          if (df$PassLength[1] > 15) {return("Post")} else {return ("Slant")}
        }
        if (df$PassLength[1] > 15 & CornerRoute(df, "SR")) return ("Corner")
        if (CurlRoute(df,"SR")) return("Curl")
        if (ComebackRoute(df,"SR")) return("Comeback")
        if (FlyRoute(df, "SR")) return("Fly")
        
      }
    } else {
      # offense direction is north
      # check on far left side of field
      if (df$y[1] > 26.65) {
        if (InDigRoute(df,"NL")) return("In/Dig")
        if (OutRoute(df,"NL")) {
          if (df$PassLength[1] > 8) {return("Out")} else {return ("Flat")} ##8 clap go bruins
          }
        if (SlantRoute(df,"NL")) {
          if (df$PassLength[1] > 15) {return("Post")} else {return ("Slant")}
        }
        if (df$PassLength[1] > 15 & CornerRoute(df, "NL")) return ("Corner")
        if (CurlRoute(df,"NL")) return("Curl")
        if (ComebackRoute(df,"NL")) return("Comeback")
        
      } else {
        # right side
        if (InDigRoute(df,"NR")) return("In/Dig")
        if (OutRoute(df,"NR")) {
          if (df$PassLength[1] > 8) {return("Out")} else {return ("Flat")} ##8 clap go bruins
          }
        if (SlantRoute(df,"NR")) {
          if (df$PassLength[1] > 15) {return("Post")} else {return ("Slant")}
        }
        if (df$PassLength[1] > 15 & CornerRoute(df, "NR")) return ("Corner")
        if (CurlRoute(df,"NR")) return("Curl")
        if (ComebackRoute(df,"NR")) return("Comeback")
        if (FlyRoute(df, "NR")) return("Fly")
        
      }
    }
    return("idk")
  }
  return(NA)
}
```


```{r}
# create new route type variable
gr.spplays1$RouteType <- NA
for (play.id in unique(gr.spplays1$playId)) {
  pl.df <- gr.spplays1[gr.spplays1$playId==play.id,]
  for (nfl.id in unique(pl.df$nflId)) {
    if (is.na(nfl.id)) {
      next
    } else {
      nfl.pl <- pl.df[pl.df$nflId==nfl.id,]
      # get subset of dataframe between ball snap and when pass is arrived
      ball.snap.ind <- which(nfl.pl$event=="ball_snap")
      pass.arr.ind <- which(nfl.pl$event=="pass_arrived")
      if (length(pass.arr.ind) != 0) {
        nfl.pl <- nfl.pl[ball.snap.ind:pass.arr.ind,]
      }
      # now assign route type based on position
      gr.spplays1[!is.na(gr.spplays1$nflId) & gr.spplays1$playId==play.id & gr.spplays1$nflId==nfl.id,]$RouteType <-
        routeHandler(nfl.pl)
    }
  }
}
```


```{r}
View(gr.spplays1)
```





Ensure they
Direction: 0-360. Goes clockwise starting from negative side of x axis.

X and y are swapping. So a player running along a yard line (left/right) is changing only y. 

For a forward slant route:
- Player's Position must be receiver
  - Bring in player data match by nflId, grab position
  
- Player's x coord must positively increase
- Player's dir should change from either 90/270 to ~120-150/300-330 respectively